
\chapter{Introduction}

\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Global Pre-Submission Formatting Checklist (remove before final):}
\begin{itemize}
  \item \textbf{Dashes:} Search for lone \texttt{-} used as a dash (should be
        \texttt{--} for ranges and \texttt{---} for em-dashes). Also check
        \texttt{--} used where \texttt{---} is needed (sentence breaks,
        parenthetical clauses).
  \item \textbf{Bold (\texttt{\textbackslash textbf}):} Confirm every bold
        instance is a defined term, table header, or finding label.
        Remove decorative bold from running prose.
  \item \textbf{Italic (\texttt{\textbackslash textit} / \texttt{\textbackslash emph}):}
        Italics are for first-use technical terms and titles only.
        Remove italic emphasis used for rhetorical stress.
  \item \textbf{Yellow boxes:} Remove all \texttt{Working Draft Note} boxes
        before submission.
  \item \textbf{Figures missing:} \texttt{fig\_sleep\_stage\_distribution.pdf}
        --- run \texttt{python generate\_distribution\_figure.py} (no data
        required, uses hardcoded counts, works on any machine).
        \texttt{fig\_confusion\_matrix.pdf} --- run
        \texttt{python generate\_missing\_figures.py} on the benchmark
        desktop (requires trained model cache).
\end{itemize}
\end{minipage}}
\end{center}

\section{Introduction and Motivation}

This thesis originated from a practical challenge encountered during a subject-independent sleep stage classification project. The task involved training machine learning models to automatically label 30-second EEG epochs as Wake, N1, N2, N3, or REM. While the classification performance was successful, the iteration speed proved to be a significant bottleneck.

The standard evaluation protocol for this domain, Leave-One-Subject-Out (LOSO) cross-validation, requires training a separate model for each subject to ensure subject-independence. With 128 subjects in the dataset, every adjustment, whether modifying a correlation threshold, varying the feature count, or switching between XGBoost and Random Forest—triggered a full retraining of 128 individual models. Because the initial pipeline lacked a mechanism for modular execution or state persistence, exploring even a modest grid of nine configurations required days of wall-clock time for computations that, logically, should have been near-instantaneous.

The central question shifted from "how can the classifier be improved?" to "why are models being retrained when their underlying configurations remain unchanged?" In a LOSO sweep, the vast majority of models remain valid if only a subset of parameters is modified. This realized computational waste was identified not as a classification failure, but as a systematic failure in caching and state management. Consequently, the focus of this research shifted from the classification task itself to the development of a caching optimization framework, using sleep staging as the primary validation vehicle for addressing redundant computation in iterative ML experimentation.

ML projects frequently fail not due to algorithmic limitations but due to infrastructure deficiencies. Rzig et al.\ report that 61.8\% of CI/CD pipeline modifications in ML projects address build policies and dependency management rather than model logic~\cite{rzig2024empirical}. This aligns with the concept of \textit{hidden technical debt} in ML systems, where the actual model code constitutes only a small fraction of the overall system~\cite{googlemlops2023}. The framework developed in this thesis aims to reduce this technical debt through automated, deterministic caching.

The broader landscape reinforces the urgency of this work. AI and ML research has expanded rapidly in recent years: AI-related publications have increased 2.3 times since 2015, now comprising nearly 42\% of all computer science publications~\cite{AIIndex2023}. Training compute for notable models has been reported to double approximately every five months, and some large-scale training runs have produced substantial CO$_2$ emissions~\cite{AIIndex2025}. Redundant computation is not limited to frontier AI: the 21-hour Random Forest sweeps that motivated this thesis are a smaller-scale instance of the same general inefficiency. As ML systems become more complex and resource-intensive, MLOps pipelines should incorporate deterministic caching to reduce avoidable recomputation wherever it occurs.

This growth highlights a critical gap in current development practices. As Kreuzberger et al.\ observe, many data science projects remain ``manual to a great extent,'' functioning as experimental Proofs of Concept that fail to operationalize efficiently~\cite{kreuzberger2023mlops}. The initial manual iteration cycle encountered in this research---where simple parameter changes triggered 21-hour re-training runs---is a textbook example of this failure mode. By automating the caching decision through a deterministic framework, this thesis moves the pipeline from a manual prototype to an operationalized workflow aligned with the principles of sustainable and reproducible AI. When experimentation relies on manual tracking rather than deterministic caching, it becomes exceedingly difficult to guarantee exact reproducibility of prior results, as unrecorded stochastic variations or configuration drift inevitably occur. This thesis therefore treats reproducibility as a first-class requirement: identical data partitions and tracked configurations must produce identical fingerprint decisions and equivalent predictive outputs.

Beyond computational efficiency, caching at the LOSO fold level preserves the complete trained model state for each held-out subject individually. Unlike aggregated cross-validation metrics---which collapse 128 fold-specific models into a single summary number---persisting the individual models enables post-hoc scientific analysis: inspecting per-fold feature importances, identifying subject-specific failure modes, or examining structural differences between models trained on different subject populations. This persistence allows domain experts and machine learning practitioners to perform interpretability analyses---such as inspecting decision tree splits or feature importances for a specific subject---without requiring expensive retraining. This ancillary benefit is not a central motivation of this thesis, but it follows directly from caching at the model level rather than the results level.

\section{Research Problem}

Despite rapid advancements in Machine Learning (ML), iterative experimentation workflows face significant computational inefficiencies, particularly in cross-validation scenarios. Leave-One-Subject-Out (LOSO) cross-validation, the standard for evaluating model generalization across individuals, requires training $N$ models for $N$ subjects. In this research, a single 9-configuration XGBoost sweep across 128 subjects required approximately 13,600 seconds ($\approx$3.8 hours) from a cold start. The overhead was significantly higher for Random Forest, where the same sweep exceeded 77,000 seconds ($\approx$21 hours). Consequently, a full 18-configuration experiment can span from several hours to multiple days per complete cold run. When researchers iterate on hyperparameters or feature selection strategies, this process is typically repeated in its entirety, leading to massive redundant computation even when the majority of trained models remain valid under the new configuration.

The scope of the experimental grid was itself constrained by this computational reality. The original project design involved five to six model families, four correlation thresholds, four top-$K$ feature counts, and 128-fold LOSO evaluation---on the order of 12{,}000 individual training operations. This thesis scopes the empirical benchmark to 18 representative configurations ($2 \times 3 \times 3$). This controlled reduction was necessary to practically execute both the cold-start (fully uncached) and warm-start (fully cached) benchmarking cycles required to accurately quantify the framework's time savings---each of which requires a complete 2{,}304-fold run. The caching framework was, in effect, developed to make the systematic evaluation of caching itself tractable.

This thesis addresses three interconnected challenges in managing these LOSO workflows:

\begin{itemize}

\item \textbf{Redundant Computation Across Experimental Iterations:}
ML experimentation requires repeatedly modifying specific components
(hyperparameters, feature subsets, model architectures) while keeping others
constant. Commercial MLOps tools such as DVC~\cite{dvc_pipelines} and ZenML provide
hierarchical pipeline caching, where downstream changes do not
invalidate upstream artifacts. However, adapting these tools to LOSO workflows
requires explicit parameterisation of the held-out subject and careful cache
key design, a process that is neither standardised nor validated in existing
literature. Without a validated framework, researchers face a choice between
no caching (prohibitive recomputation costs) or manual cache management
(error-prone and difficult to reproduce).

\item \textbf{Ensuring Cache Validity Across Cross-Validation Folds:}
In LOSO cross-validation, each fold's trained model is valid only for
evaluating its specific held-out subject. A model trained with subject $S_i$
held out produces scientifically invalid predictions when applied to evaluate
subject $S_j$ held out --- not due to data leakage (the train/test split is
maintained) but due to applying cached results to an incorrect experimental
condition.

While general-purpose caching systems like DVC can be configured to handle this
by parameterising the held-out subject, there is no validated framework
demonstrating best practices for LOSO cache key design. Without explicit
guidance, researchers risk either overly conservative cache invalidation
(recomputing valid results) or incorrect parameterisation (potential cache
validity violations). This thesis established a validated approach where
held-out subject identifiers are systematically included in fingerprint-based
cache keys.

\item \textbf{Unified Fingerprinting for Reproducibility and Cache Validity:}
Existing MLOps tools separate experiment tracking (MLflow, W\&B) from
pipeline caching (DVC, ZenML), requiring researchers to manually coordinate
logged configurations with cached artifacts~\cite{zaharia2018mlflow,%
wandb2024docs,dvc_pipelines}. This separation becomes particularly
challenging in LOSO scenarios where fold identity must be consistently tracked
across both systems. This thesis implemented cryptographic fingerprinting
(SHA-256 of canonical JSON configurations) as a unified solution: a single
fingerprint simultaneously ensures cache validity, reproducibility, and
experiment identification. The approach was validated by demonstrating
deterministic fingerprint generation across 18 experimental configurations
and 128 cross-validation folds, with zero fingerprint collisions and
complete result reproducibility.

\end{itemize}

This thesis addresses these challenges through a fingerprint-based caching
framework for LOSO cross-validation experiments. The framework uses SHA-256
hashing of canonical JSON configuration representations to generate deterministic
cache keys, with held-out subject identifiers included to ensure cache validity
across folds. The approach was validated through a sleep stage classification
case study using 128 subjects from the BOAS (Bitbrain Open Access Sleep)
dataset, measuring cache hit rates, computational time savings, storage
requirements, and reproducibility across experimental iterations.


\section{Research Questions}

\textbf{Main Research Question:}

How can fingerprint-based caching reduce computational costs in
Leave-One-Subject-Out cross-validation while ensuring reproducibility
and preventing data leakage?

\begin{itemize}
\item \textbf{RQ1 (Efficiency):} What time savings and cache hit rates were
achieved through fingerprint-based caching of extracted features and trained
models in LOSO cross-validation?
\item \textbf{RQ2 (Reproducibility):} How does SHA-256 configuration
fingerprinting ensure consistent and reproducible experiment results across
multiple runs?
\item \textbf{RQ3 (Scalability):} How does caching performance scale with
dataset size, from 10 subjects to the full 128-subject dataset?
\item \textbf{RQ4 (Trade-offs):} What are the storage costs versus
computational savings of the caching approach, and under what conditions
does caching remain viable?
\end{itemize}



\section{Research Design and Methodology}

This thesis follows a quantitative experimental methodology to evaluate a fingerprint-based caching framework in an EEG sleep staging pipeline. The approach combines implementation, controlled benchmarking, and validation-oriented follow-up experiments.

\subsection{Methodological Phases}

\begin{enumerate}
    \item \textbf{Framework Implementation:}
    A deterministic fingerprinting mechanism based on SHA-256 hashing of canonical experiment configuration is implemented together with a two-level cache architecture:
    (i) feature-level caching and (ii) LOSO model-level caching.
    The held-out subject identifier is included in the model cache key to prevent cross-fold reuse.

    \item \textbf{Core Benchmark Evaluation:}
    A full factorial benchmark is executed for two model families (XGBoost, Random Forest) across
    three correlation-threshold settings and three feature-count settings (18 configurations total),
    each under 128-fold Leave-One-Subject-Out (LOSO) evaluation.
    For each configuration, cold-start and warm-cache runs are compared using:
    total runtime, cache-loading time, speedup ratio, storage cost, and classification performance.

    \item \textbf{Extended Validation:}
    Additional experiments assess generalizability and robustness:
    multi-model cache viability analysis, scaling experiments with increasing subject counts, and feature-selection efficiency analysis (ANOVA versus Mutual Information).


\end{enumerate}

\subsection{Evaluation Dimensions}

The evaluation is structured along two orthogonal dimensions:

\begin{itemize}
    \item \textbf{Optimization Objective (WHY):} 
    Quantifying the reduction in computational latency, resource efficiency, and cumulative time savings achieved through deterministic cache reuse across iterative experimental cycles.

    \item \textbf{Caching Placement (WHERE):} 
    Assessing the complementary effects of a two-tier persistence strategy designed to optimize different stages of the ML pipeline:
    \begin{itemize}
        \item \textbf{Tier 1: Persistent Feature Store.} One-time extraction and
              persistent storage of 149-dimensional feature vectors from
              6-channel EEG signal processing. Features are extracted once per
              subject and stored in a shared, experiment-agnostic cache. This
              tier decouples expensive signal processing from downstream model
              training: once computed, features are reused across all 18
              experimental configurations without recomputation.
        \item \textbf{Tier 2: Fingerprint-based Model Cache.}
              Configuration-aware training and storage of fold-specific models.
              For each LOSO fold, this tier retrieves cached features from
              Tier~1, optionally applies feature selection (correlation
              thresholding and top-$K$ ranking), trains a model, and stores
              it under a SHA-256 fingerprint that encodes the full
              configuration including the held-out subject identifier.
              This prevents redundant retraining when only hyperparameters or
              feature-selection parameters change across experimental iterations.
    \end{itemize}
\end{itemize}

\subsection{Validity and Reproducibility Criteria}

A cache lookup is considered valid only when the fingerprinted configuration matches.
If any tracked parameter changes, the fingerprint changes and cached outputs are bypassed.
Empirical validation focuses on two criteria:
(1) performance gains from warm-cache execution compared with cold retraining, and
(2) consistency of predictive outputs between cached and freshly trained models in the reported benchmark suite.

\section{Structure of the Thesis}
This thesis is structured as follows:

\begin{itemize}
\item \textbf{Chapter 1: Introduction} presents the motivation rooted in
practical LOSO runtime problems, the research problem, research questions,
and research approach.

\item \textbf{Chapter 2: Background} covers ML research methodology,
computational efficiency challenges, sleep medicine fundamentals, EEG signal
processing basics, ML ensemble methods, and configuration fingerprinting
and caching technologies.

\item \textbf{Chapter 3: Related Works} reviews existing MLOps tools
(MLflow, DVC, ZenML), sleep stage classification methods grouped by technique,
and identifies the gap in LOSO-aware caching that this work addresses.

\item \textbf{Chapter 4: Methodology and Results} details the fingerprint-based
caching framework, the BOAS dataset, preprocessing pipeline, extraction of
149 EEG features per epoch, the LOSO cross-validation experimental design,
and the full experimental results including cache speedups, classification
performance, model viability analysis, and discussion.

\item \textbf{Chapter 5: Summary and Future Work} summarises the contributions
of the fingerprint-based caching framework and discusses extensions including
two-tier RAM+Disk caching and cross-experiment cache reuse.
\end{itemize}
