\chapter{Background}

\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Working Draft Note:} 
Lenghth goal 8 pages currently reached  including yellow to does
\end{minipage}}
\end{center}
\section{Domain: Machine Learning Research Methodology}
The landscape of Machine Learning (ML) research and development is
characterized by rapid iteration and experimentation. According to the
AI Index Report 2023, AI-related publications have more than doubled since
2015, reflecting the field's exponential growth~\cite{AIIndex2023}. Data
scientists and ML engineers continuously explore different model architectures,
algorithms, hyperparameters, and data preprocessing techniques to achieve
optimal performance. This iterative process, while essential for innovation,
introduces significant challenges concerning efficiency and reproducibility.
A single experiment may require hours or days of computation, and researchers
often repeat similar computations across multiple runs. The cumulative effect
translates into substantial costs in GPU hours, cloud computing expenses,
and researcher productivity~\cite{schubotz2022caching}.

To manage this complexity, standardized process models have emerged.
CRISP-DM (Cross-Industry Standard Process for Data Mining), developed in
1996, remains widely adopted~\cite{crispdm2000}. However, its limitations
for modern ML---particularly lack of emphasis on model monitoring and
maintenance---led to CRISP-ML(Q), which extends CRISP-DM with quality assurance
methodology tailored for ML applications~\cite{studer2021crispmlq}. This
framework aligns with modern MLOps principles, which emphasize automation,
continuous integration/deployment (CI/CD), and reproducibility throughout
the ML lifecycle~\cite{kreuzberger2023mlops}.

Tools such as MLflow~\cite{zaharia2018mlflow} and MLXP~\cite{arbel2024mlxp}
support experiment tracking by recording parameters, metrics, code versions,
and artifacts for each run, enabling comparison and reproducibility. Despite
these advances, reproducibility remains a persistent challenge---particularly
in subject-independent evaluation workflows such as LOSO cross-validation,
where experiment tracking tools record parameters globally but do not natively
encode fold identity into cache keys. Without explicit fold-aware engineering,
a cached model trained with Subject~$S_i$ held out could silently be reused
for Subject~$S_j$, violating the experimental
protocol~\cite{semmelrock2025reproducibility}. As Kreuzberger et al.\ define
it, reproducibility is the ``ability to obtain the exact same results'' given
identical inputs, code, and execution environment (Principle~P3)~\cite{kreuzberger2023mlops},
a goal that depends on three controllable factors~\cite{schubotz2022caching}:
\begin{itemize}
    \item \textbf{Lack of Records and Hyperparameter Inconsistency:} 
          Unrecorded decisions or parameter changes hinder replication.
    
    \item \textbf{Data and Environment Volatility:} Dataset alterations, 
          framework updates, and inconsistent compute environments cause 
          divergent outcomes. Empirical analysis shows that 61.8\% of 
          CI/CD pipeline modifications in ML projects are driven by 
          such dependency issues~\cite{rzig2024empirical}.

    \item \textbf{Inherent Randomness:} Machine learning algorithms, 
          particularly Feedforward Neural Networks (FFNNs), exhibit 
          significant variation due to stochastic model initialization 
          and training~\cite{raste2022quantifying}. These effects are 
          further amplified by hardware-level non-determinism; for 
          instance, NVIDIA's cuDNN routines utilize atomic operations 
          that introduce "truly random floating point rounding errors" 
          during execution~\cite{nvidia2024cudnn}.
\end{itemize}


These challenges are particularly acute during iterative experimentation,
where researchers may run hundreds of configurations before identifying
promising approaches. Without intelligent management of computational
artifacts, substantial resources are wasted on redundant calculations.
This observation motivates the central contribution of this thesis: a
fingerprint-based caching framework that persists and reuses computational
artifacts (extracted features and trained models) across iterative modeling
phases while maintaining reproducibility through cryptographic configuration
fingerprinting. Kreuzberger et al.'s Principle~P7 requires metadata tracking
for maintaining model lineage~\cite{kreuzberger2023mlops}; the fingerprint-based
framework implements this directly by hashing a canonical JSON representation
of the complete experiment state---data configuration, code version, model
parameters, and held-out subject identifier---into a SHA-256 fingerprint that
simultaneously automates lineage tracking and serves as a deterministic cache key.

\section{Context: Computational Efficiency in Research}

Building on the iterative experimentation challenges described in
Section 2.1, this section examines the computational efficiency bottlenecks
that motivate intelligent caching frameworks.

The increasing complexity and scale of Machine Learning models have
substantially raised computational demands. Training models, tuning
hyperparameters, and running repeated experiments consume vast compute,
memory, and time resources, creating bottlenecks in research productivity
and driving up costs.

Several factors drive this resource consumption:

\begin{itemize}
\item \textbf{Large Datasets:} Massive datasets require intensive I/O and
      memory usage.

\item \textbf{Complex Models:} Both deep neural networks and ensemble
      methods (e.g., gradient boosting with thousands of trees, large
      random forests) require substantial computational resources for
      training and inference.

\item \textbf{Iterative Experimentation:} Frequent experiments often repeat
      computations unnecessarily when modifying a subset of parameters while
      keeping others constant.

\item \textbf{Hyperparameter Optimization:} Searching large hyperparameter
      spaces is inherently compute-intensive.

\item \textbf{Rigorous Validation:} Cross-validation strategies (e.g.,
      k-fold, leave-one-out) multiply training costs by the number of folds.
\end{itemize}

While caching can help, hardware-level research in cache management---such as the ML-aided strategies surveyed by Krishna (2025)---focuses primarily on optimizing low-level latency and hardware replacement policies~\cite{krishna2025cache}. These systems are designed to predict microsecond-scale access patterns rather than addressing macro-level requirements such as matching high-level experiment configurations to saved artifacts. For EEG research, such low-level caching does not prevent the redundant recomputation of 128-fold cross-validation when hyperparameters change.

This thesis addresses these inefficiencies through a deterministic, fingerprint-based caching framework. By encoding complete experiment configurations (including cross-validation fold identity) into cryptographic SHA-256 cache keys, the framework enables automatic result reuse that is mathematically tied to the scientific parameters, ensuring both efficiency and reproducibility.

\section{Technology: Configuration Fingerprinting and Caching}

Efficient ML experimentation depends on configuration management and intelligent
result reuse. This section reviews technologies directly relevant to the
fingerprint-based LOSO caching approach developed in this thesis.

A \textbf{cache} is a temporary store of computed results that can be reused when the same computation is requested again; its defining property is a \textit{validity condition}---a cached result is only reusable if the inputs that produced it remain unchanged. \textbf{Persistent storage}, by contrast, refers to the durable storage of artifacts whose validity does not depend on experimental configuration, such as raw data or fixed intermediate representations. Schubotz et al.\ (2022) formalize this distinction in data science experiments, demonstrating that reproducible result reuse requires explicit tracking of what produced each artifact and under which conditions it remains valid~\cite{schubotz2022caching}. This thesis operationalizes both concepts: extracted EEG features constitute persistent storage (computed once per dataset regardless of model choices), while trained LOSO models constitute a configuration-dependent cache (valid only for their specific fingerprint, as detailed in Section~\ref{sec:storage_vs_caching}).

The adoption of Leave-One-Subject-Out (LOSO) cross-validation is driven by the requirement for algorithmic stability in high-dimensional EEG datasets. As established by Kearns and Ron (1999), leave-one-out estimators provide essential ``sanity-check'' bounds for generalisation error, ensuring that the model does not merely overfit to subject-specific artefacts~\cite{kearns1999}. However, the computational overhead of LOSO remains a significant barrier to iterative research. Jayaram (2018) emphasises that for EEG-based systems to move beyond the laboratory, processing pipelines must be validated across diverse datasets without requiring constant, redundant retraining~\cite{jayaram2018}. In this thesis, LOSO is utilised not only as a validation metric but as a primary driver for the caching framework: the repetitive nature of the 2{,}304 required training runs provides a high-volume workload where deterministic fingerprinting can demonstrate maximum efficiency gains.

\subsection{Configuration Fingerprinting}

Configuration fingerprinting uses cryptographic hash functions (e.g., SHA-256) 
to create unique identifiers for experiment configurations. By hashing a 
canonical representation of all parameters, these fingerprints enable 
deterministic identification of identical configurations and automatic 
cache invalidation when parameters change.

Modern MLOps tools employ variants of this approach:
\begin{itemize}
    \item \textbf{DVC} logs pipeline signatures combining dependency contents, 
          parameters, and commands~\cite{dvc_pipelines}.
    \item \textbf{MLflow} tracks experiment metadata to enable versioning and 
          visual comparison~\cite{zaharia2018mlflow}.
    \item \textbf{MLXP} enforces reproducibility by linking Hydra-based 
          configurations with specific Git commit hashes~\cite{arbel2024mlxp}.
\end{itemize}

\subsection{Hash-Based Caching in MLOps}

MLOps pipeline caching implements content-addressable storage: when pipeline
inputs and code match previous runs (verified through hash comparison),
cached outputs are reused~\cite{dvc_pipelines}. DVC's run cache,
for example, combines dependency hashes with parameter configurations to
create cache keys. When a match is found, execution is skipped and previous
results are restored.

In practice, content-addressable caches are typically implemented as
filesystem-based key--value stores: artifacts are saved to files whose names
are derived from the configuration fingerprint, and a cache lookup is simply
a file-existence check using the hash-derived filename. This is distinct from
an in-memory hash table; lookup performance is governed by the operating
system's directory indexing rather than constant-time hashing.

This approach provides reproducibility: identical configurations produce
identical cache keys, ensuring cached results correspond to recorded
parameters~\cite{schubotz2022caching}.

Google's MLOps maturity model~\cite{googlemlops2023} describes a progression
from Level~0 (fully manual processes) to Level~1 (automated ML pipelines) in
which experiment metadata, artifact versioning, and reproducible pipeline
execution are central requirements. Hash-based caching supports this
progression by providing the technical substrate for traceability: only through
unique configuration fingerprints can the exact data and code versions that
produced a specific artifact be identified with certainty.

\subsection{Related Caching Approaches}

Caching is applied across many computing domains, but with fundamentally
different objectives. LLM inference caches manage attention key-value tensors
to reduce autoregressive latency~\cite{kwon2023pagedattention}; hardware-level
systems optimize replacement policies for microsecond-scale memory
access~\cite{krishna2025cache}; and large-scale scientific computing uses
distributed file caches for petabyte data distribution~\cite{sim2023xcache}.
In contrast to these runtime or hardware-oriented approaches, this thesis
implements \textit{experiment-level} caching: artifacts are indexed by the
complete scientific configuration that produced them, and validity depends on
configuration identity and cross-validation fold identity rather than access
frequency or latency.

\subsection{Reproducibility Considerations}

Despite fingerprinting, perfect reproducibility remains challenging. Sources
of non-determinism include random seed variation, software version differences,
hardware-specific floating-point operations, and non-deterministic GPU
algorithms~\cite{raste2022quantifying,nvidia2024cudnn}.

Bouthillier et al.\ (2021) demonstrate that variance from data sampling,
parameter initialization, and hyperparameter choice markedly impacts benchmark
results across deep learning architectures, and that these sources of variation
are routinely underestimated in published
comparisons~\cite{bouthillier2021variance}. Raste et al.\ (2022) quantify
inherent randomness specifically in tree-based methods (Random Forests, GBMs)
and Feedforward Neural Networks, finding that variation from data splitting
exceeds variation from model training, while FFNNs exhibit the largest
training-induced variance~\cite{raste2022quantifying}. NVIDIA's cuDNN
documentation explicitly states that certain GPU reduction operations use
atomic operations that introduce non-deterministic floating-point rounding
errors~\cite{nvidia2024cudnn}.

Configuration fingerprinting addresses the controllable factors (seeds,
hyperparameters, data splits, software versions) while accepting
hardware-induced variance as irreducible. For caching purposes, cached
results are considered valid if fingerprints match, with the understanding
that minor numerical differences may occur when reproducing results across
different hardware environments.

% ============================================================================
\section{Sleep Medicine Fundamentals}
\label{sec:sleep_medicine}
% ============================================================================

The caching framework is domain-agnostic, but its validation in this thesis
is grounded in a specific application: automatic sleep stage classification
from EEG signals. The following sections introduce the sleep medicine domain
and EEG signal processing concepts that define the data, labels, and
feature space used in the experiments.

\subsection{Sleep Stages and Their Characteristics}

The American Academy of Sleep Medicine (AASM) defines five sleep stages,
scored in 30-second epochs from polysomnographic recordings~\cite{berry2017aasm}.
These stage labels constitute the classification targets ($y$-variables) in this
thesis:

\begin{itemize}
\item \textbf{Wake (W):} Low-amplitude, high-frequency EEG; alpha waves
      (8--13~Hz) predominate during relaxed wakefulness.

\item \textbf{N1 (Light Sleep):} Mixed-frequency, low-amplitude activity.
      The most difficult stage to classify automatically due to its
      transitional nature (typically 1--5 minutes).

\item \textbf{N2 (Sleep):} Comprises 45--55\% of total sleep time. Defined
      by sleep spindles (12--16~Hz as used in this thesis's feature extractor;
      AASM defines spindles at 11--16~Hz) and K-complexes.

\item \textbf{N3 (Deep Sleep):} Dominated by high-amplitude delta waves
      (0.5--4~Hz, $\geq$20\% of epoch). Most prevalent in the first third
      of the night.

\item \textbf{REM:} Mixed-frequency low-amplitude EEG similar to wakefulness,
      with rapid eye movements and skeletal muscle atonia.
\end{itemize}

\subsection{EEG Acquisition and the Need for Automation}

The raw input data consists of 6-channel EEG signals recorded via the
international 10-20 electrode placement system~\cite{niedermeyer2011}.
Labels indicate brain region (F=frontal, C=central, O=occipital) and
hemisphere (odd=left, even=right). This thesis uses channels F3, F4, C3,
C4, O1, O2, providing bilateral coverage of frontal, sensorimotor, and
visual cortices. In the context of the caching architecture, these channel
identifiers and the 30-second epoch duration are treated as immutable
metadata in the SHA-256 fingerprint.

Manual sleep scoring by certified technologists requires approximately
1.5--2 hours per recording. For large cohorts such as the BOAS dataset
(128 subjects), manual annotation becomes prohibitively expensive,
motivating automated classification. Inter-rater agreement among human
experts is typically 80--90\%, providing a natural performance ceiling
for automated systems.

% ============================================================================
\section{EEG Signal Processing Basics}
\label{sec:eeg_processing}
% ============================================================================

EEG signals are non-stationary time series whose classification relies on
identifying characteristic oscillatory patterns in specific frequency
bands~\cite{niedermeyer2011}. These frequency bands form the basis of the
feature extraction pipeline and, consequently, a core parameter in the
cache fingerprint.

\subsection{Frequency Bands}

In this work, EEG signals are decomposed into six frequency bands.
Table~\ref{tab:eeg_bands} lists the ranges as implemented in the feature
extraction pipeline. Two boundaries deviate from the conventional literature
values~\cite{niedermeyer2011}: Beta starts at 16~Hz rather than the
conventional 13~Hz to avoid overlap with the Sigma band (12--16~Hz), and
Gamma is capped at 40~Hz to match the bandpass filter cutoff rather than
the commonly cited 45~Hz upper limit.

\begin{table}[htbp]
\centering
\caption[EEG frequency bands as used in this thesis]{EEG frequency bands
as implemented in this thesis and their sleep-stage associations.
Beta and Gamma boundaries differ slightly from conventional ranges
(see text).}
\label{tab:eeg_bands}
\begin{tabular}{lll}
\toprule
\textbf{Band} & \textbf{Range (Hz)} & \textbf{Sleep Stage Association} \\
\midrule
Delta    & 0.5--4.0   & Predominant in N3 (deep sleep); indicates sleep pressure \\
Theta    & 4.0--8.0   & Associated with N1 and REM; drowsiness marker \\
Alpha    & 8.0--13.0  & Relaxed wakefulness; key marker for Stage W \\
Sigma    & 12.0--16.0 & Sleep spindles; definitive marker for N2 sleep \\
Beta     & 16.0--30.0 & Active wakefulness; power decreases during NREM \\
Gamma    & 30.0--40.0 & High-frequency processing; typically minimal in sleep \\
\bottomrule
\end{tabular}
\end{table}

Power spectral density (PSD) in these bands, computed via Welch's method, forms the core of the frequency-domain feature extraction pipeline. By quantifying the relative power in these specific ranges, the XGBoost and Random Forest models can distinguish between the synchronized, high-amplitude waves of deep sleep and the desynchronized, low-amplitude patterns of wakefulness.

\subsection{Standard Preprocessing Pipeline}

Raw EEG signals contain multiple sources of contamination that must be
addressed before feature extraction. Standard EEG preprocessing targets
the following noise sources:

\begin{itemize}
\item \textbf{DC offset and slow drift:} Removed by high-pass filtering
      (typically $\geq$0.5~Hz).
\item \textbf{High-frequency noise:} Removed by low-pass filtering
      (typically $\leq$40~Hz for sleep staging).
\item \textbf{Power line interference:} A 50~Hz (Europe) or 60~Hz (USA)
      notch filter removes electrical mains noise.
\item \textbf{Downsampling:} High sampling rates are reduced to the minimum
      sufficient to capture all sleep-relevant frequencies ($\leq$40~Hz),
      halving computational costs while preserving signal fidelity.
\item \textbf{Epoch rejection:} Epochs contaminated by motion artefacts,
      excessive amplitude, or flat channels may be excluded before feature
      extraction; whether to apply this step depends on dataset quality.
\end{itemize}

These steps constitute the preprocessing pipeline used in this thesis,
producing clean 30-second epochs ready for feature extraction. The pipeline
was originally developed for a prior subject-independent sleep staging project
and is adopted here with minor adaptations; this thesis contributes the
fingerprint-based caching layer on top of it, treating the preprocessing
as a fixed, validated substrate.

% ============================================================================
\section{Machine Learning Ensemble Methods}
\label{sec:ensemble_methods}
% ============================================================================
Ensemble methods combine multiple learners to produce a stronger predictor
than any individual model~\cite{hastie2009elements}. They are particularly
effective for tabular classification tasks, which motivates their use in
this thesis. Two ensemble strategies differ fundamentally in how they
aggregate learners: \textit{bagging} trains learners independently in
parallel (Random Forest), while \textit{boosting} trains them sequentially,
each correcting the errors of the previous (XGBoost). This section covers
both families.

\subsection{Decision Trees}

A decision tree recursively partitions the feature space by selecting
feature thresholds that maximise class separation (measured by Gini
impurity or information gain). Individual trees are fast to train but
prone to overfitting. Ensemble methods address this by combining many
trees trained under different conditions.

\subsection{Bagging: Random Forest}

Random Forest~\cite{breiman2001random} trains each tree on a bootstrap
sample of the training data and restricts each split to a random subset
of features. The final prediction is the majority vote across all trees,
reducing variance without substantially increasing bias.

Random Forest typically achieves strong out-of-the-box performance on
tabular data but produces large serialized model files (600--1{,}544~MB
per fold for 149 features and 127 training subjects). This storage
overhead is a key factor in its caching viability, as discussed in
Chapter~4.

\subsection{Boosting: XGBoost}

XGBoost (Extreme Gradient Boosting)~\cite{chen2016xgboost} extends
gradient boosting with
regularisation terms, sparsity-aware splitting, and cache-efficient tree
construction. Unlike Random Forest, boosting models use shallow trees
(depth 3--6) with aggressive regularisation, producing compact serialized
representations of 1--2~MB per fold---making them highly cache-viable.

\subsection{Why These Models for EEG Feature Classification}

Random Forest and XGBoost are selected for this thesis because they
represent two contrasting ensemble strategies (bagging vs.\ boosting)
with fundamentally different serialization characteristics. This contrast
directly serves the caching evaluation: XGBoost produces compact models
that are fast to cache and load, while Random Forest produces large models
that stress-test the caching framework's viability boundaries.
Additionally, both models handle the 149-dimensional feature space
effectively and provide interpretable feature importance scores for
post-hoc analysis.

% ============================================================================
\section{Cross-Validation in Subject-Independent ML}
\label{sec:loso_background}
% ============================================================================

EEG signals exhibit high inter-subject variability: brain activity patterns
differ substantially between individuals due to anatomical differences,
electrode placement tolerances, and inter-individual variation in
neurophysiology~\cite{huang2023intersubject}. This variability means that
a model trained on data from subjects $S_1, \ldots, S_{N-1}$ and evaluated
on a held-in portion of those same subjects' data will overestimate real-world
performance on a new, unseen subject. Standard $k$-fold cross-validation does
not account for this: train and test folds typically contain epochs from the
same subjects, producing accuracy estimates that are misleadingly optimistic
for subject-independent deployment.

Leave-One-Subject-Out (LOSO) cross-validation addresses this by training
on all subjects except one and evaluating on the excluded subject, cycling
through every subject as the test case. Kunjan et al.\ demonstrate directly
that for scalp-recorded EEG classification, $k$-fold results are misleading
for unseen-subject performance, and conclude that LOSO evaluation is
necessary for any EEG-based machine learning study~\cite{kunjan2021loso}.

The computational cost is proportional to the cohort size: with $N = 128$
subjects, a single experimental configuration requires 128 separate model
training runs. Each run produces a distinct trained model valid only for
evaluating its specific held-out subject --- a model trained with subject
$S_i$ excluded cannot be reused to evaluate subject $S_j$, because the
training population differs. This fold-specificity is what makes LOSO the
primary driver of the caching framework: without persistent, fold-aware
storage of trained models, every configuration change triggers 128
redundant retraining operations regardless of how little the configuration
actually changed.

% ============================================================================
\section{Machine Learning Research Methodology and Computational Efficiency}
\label{sec:mlresearch_efficiency}
\label{sec:mlresearch}
\label{sec:comp_efficiency}

ML experimentation is inherently iterative: researchers explore model
architectures, hyperparameters, and preprocessing configurations across many
runs, frequently repeating computations when only a subset of parameters
changes. A single experiment may require hours or days of computation, and
the cumulative cost of redundant retraining translates into substantial GPU
hours and researcher time~\cite{schubotz2022caching}.

Standardised process models describe this iteration. CRISP-DM~\cite{crispdm2000}
defines a cyclic workflow (Business Understanding $\to$ Data Understanding
$\to$ Data Preparation $\to$ Modelling $\to$ Evaluation $\to$ Deployment)
that practitioners revisit repeatedly as findings emerge. Each cycle back
to the Modelling phase can trigger full retraining if intermediate artefacts
are not persisted. The fingerprint-based caching framework developed in this
thesis operates precisely at this transition: by memoising trained models
under a cryptographic key, it collapses repeated Modelling cycles from
hours to seconds without changing the workflow structure.

Tools such as MLflow~\cite{zaharia2018mlflow} and MLXP~\cite{arbel2024mlxp}
support experiment tracking by recording parameters, metrics, implementation
versions, and artefacts for each run, enabling comparison and reproducibility.
These tools were designed for single train/test splits and tabular grid
searches. In Leave-One-Subject-Out cross-validation with 128 subjects, a
single parameter change invalidates 128 individually trained models rather
than one---making the absence of fold-aware caching not merely inconvenient
but a potential source of silent scientific errors.

Reproducibility remains a pervasive challenge in ML
research~\cite{semmelrock2025reproducibility}. Kreuzberger et al.\ define
reproducibility as the ability to obtain identical results under identical
conditions (Principle~P3) and emphasise metadata tracking for model lineage
(Principle~P7)~\cite{kreuzberger2023mlops}. Achieving P3 requires controlling
three interdependent components: input data, tracked parameters, and the
execution environment. Challenges arise from:
\begin{itemize}
    \item \textbf{Lack of Records and Hyperparameter Inconsistency:}
          Unrecorded decisions or parameter changes hinder replication.
    \item \textbf{Data and Environment Volatility:} Dataset alterations,
          framework updates, and inconsistent compute environments cause
          divergent outcomes. Rzig et al.\ find that 61.8\% of CI/CD
          pipeline modifications in ML projects address build policies and
          dependency management rather than model logic~\cite{rzig2024empirical}.
    \item \textbf{Inherent Randomness:} Stochastic initialisation and
          hardware-level non-determinism (e.g.\ NVIDIA cuDNN atomic
          operations~\cite{nvidia2024cudnn}) affect deep learning models.
          The classical ensemble models used in this thesis (sklearn,
          XGBoost, fixed \texttt{random\_state=42}) are deterministic by
          construction and unaffected by these concerns.
\end{itemize}

The increasing complexity and scale of Machine Learning models have
substantially raised computational demands. Training models, tuning
hyperparameters, and running repeated experiments consume vast compute,
memory, and time resources. Several factors drive this resource consumption:

\begin{itemize}
\item \textbf{Large Datasets:} Massive datasets require intensive I/O and
      memory usage.

\item \textbf{Complex Models:} Both deep neural networks and ensemble
      methods (e.g., gradient boosting with thousands of trees, large
      random forests) require substantial computational resources for
      training and inference.

\item \textbf{Iterative Experimentation:} Frequent experiments often repeat
      computations unnecessarily when modifying a subset of parameters while
      keeping others constant.

\item \textbf{Hyperparameter Optimization:} Searching large hyperparameter
      spaces is inherently compute-intensive.

\item \textbf{Rigorous Validation:} Cross-validation strategies (e.g.,
      k-fold, leave-one-out) multiply training costs by the number of folds.
\end{itemize}

While caching can help, traditional research in cache management---such as the ML-aided hardware strategies surveyed by Krishna (2025)---focuses primarily on optimizing low-level latency and hardware replacement policies~\cite{krishna2025cache}. These systems are designed to predict microsecond-level access patterns rather than addressing the macro-level requirement of matching high-level experiment configurations to saved artefacts. For EEG research, such low-level caching does not prevent the redundant recomputation of 128-fold cross-validation when hyperparameters change.

Caching shifts the bottleneck from computation time to storage space --- a
trade-off whose viability depends on the serialised model size, the expected
number of re-runs, and available storage capacity. For compact models re-run
frequently, caching is clearly beneficial; for very large models run only
once, the storage overhead may outweigh the savings. This thesis addresses
this trade-off explicitly through a model viability analysis in
Chapter~\ref{chap:methodology}.

% ============================================================================
\section{Technology: Configuration Fingerprinting and Caching}
\label{sec:fingerprinting}
% ============================================================================

Efficient ML experimentation depends on configuration management and intelligent
result reuse. This section reviews technologies directly relevant to the
fingerprint-based LOSO caching approach developed in this thesis.

As established in Section~\ref{sec:loso_background}, LOSO cross-validation
is necessary for EEG-based ML to avoid inflated generalisation estimates.
Its computational cost --- 128 training runs per configuration --- makes it
the primary driver of the caching framework: 2{,}304 required training
operations provide a high-volume workload where deterministic fingerprinting
delivers maximum efficiency gains.

\begin{center}
\colorbox{yellow!20}{\begin{minipage}{0.9\textwidth}
\textbf{Note:} A prior version of this section cited Kearns \& Ron (1999)
to justify LOSO. That paper establishes PAC-learning generalisation bounds
for leave-one-out estimators --- a theoretical computer science result,
not an EEG workflow recommendation. The justification above replaces it
with the direct reference to Section~\ref{sec:loso_background} where LOSO
is properly motivated via EEG inter-subject variability (Kunjan et al.).
Verify that this framing is sufficient for the examiner.
\end{minipage}}
\end{center}

\subsection{Configuration Fingerprinting}

Configuration fingerprinting uses cryptographic hash functions (e.g., SHA-256) 
to create unique identifiers for experiment configurations. By hashing a 
canonical representation of all parameters, these fingerprints enable 
deterministic identification of identical configurations and automatic 
cache invalidation when parameters change.

Modern MLOps tools employ variants of this approach:
\begin{itemize}
    \item \textbf{DVC} logs pipeline signatures combining dependency contents, 
          parameters, and commands~\cite{dvc_pipelines}.
    \item \textbf{MLflow} tracks experiment metadata to enable versioning and 
          visual comparison~\cite{zaharia2018mlflow}.
    \item \textbf{MLXP} enforces reproducibility by linking Hydra-based 
          configurations with specific Git commit hashes~\cite{arbel2024mlxp}.
\end{itemize}

\subsection{Hash-Based Caching in MLOps}

A terminological note is warranted before proceeding. In traditional computer
science, a \textit{cache} is temporary and subject to eviction (e.g., CPU
L1/L2 caches, browser caches). In the context of MLOps and this thesis,
\textit{caching} refers to the persistent memoisation of computational
artefacts to disk---a permanent store keyed by a cryptographic fingerprint
of the configuration that produced them. These artefacts are not evicted;
they persist until explicitly deleted. The distinction within this thesis is:
the Fingerprint-based Model Cache (Tier~2) behaves as a true cache in that
a fingerprint mismatch bypasses the stored result; the Persistent Feature
Store (Tier~1) is a permanent store that is never invalidated by design.
What elevates any persistent memoisation above cold file storage is the
automated check-and-bypass protocol: the pipeline queries for a key before
computing, and loads the stored result if found rather than recomputing.

MLOps pipeline caching implements content-addressable storage: when pipeline
inputs and tracked parameters match previous runs (verified through hash comparison),
cached outputs are reused~\cite{dvc_pipelines}. DVC's run cache,
for example, combines dependency hashes with parameter configurations to
create cache keys. When a match is found, execution is skipped and previous
results are restored.

This approach provides reproducibility: identical configurations produce
identical cache keys, ensuring cached results correspond to recorded
parameters~\cite{schubotz2022caching}.

In the context of Google's MLOps maturity model~\cite{googlemlops2023}, automated pipeline execution---including artefact reuse---is a key component for progressing from Level~0 (manual processes) to Level~1 (automated training pipelines). It serves not only as an efficiency mechanism but as the technical implementation of \textit{traceability}: only through unique configuration fingerprints can the exact data and parameter versions that produced a specific artefact be identified with certainty. Fingerprinting addresses the controllable sources of irreproducibility---seeds, hyperparameters, data splits, implementation versions---while for the CPU-based sklearn and XGBoost models used in this thesis (fixed \texttt{random\_state=42}), deterministic outputs are guaranteed by construction.
