
\chapter{Related Works}


This chapter reviews existing solutions for ML experiment caching and sleep
stage classification, and identifies the specific gap this thesis addresses:
the lack of validated best practices for caching in Leave-One-Subject-Out
(LOSO) cross-validation scenarios.

% ============================================================================
\section{Academic Research on ML Caching}
% ============================================================================

Recent research explores the intersection of machine learning and cache management, though primarily from the perspective of \textit{ML-for-System-Optimization}. Krishna (2025) provides a comprehensive review of 'intelligent' caching techniques, including LSTM-based usage prediction and reinforcement learning for adaptive replacement policies~\cite{krishna2025cache}. These approaches focus on \textit{learned} caching decisions to improve hardware throughput.
\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Working Draft Note:} 
fix next cite
\end{minipage}
}
\end{center}
In contrast, the framework developed in this thesis focuses on \textit{Caching-for-ML-Workflows}. Rather than using predictive models to guess what to cache, this work employs a deterministic, content-addressed approach. This ensures that cache hits are strictly limited to identical experimental configurations—a requirement for scientific validity that the predictive models surveyed by Krishna (2025) are not designed to satisfy.

In the context of reproducibility, Schubotz et al.\ (2022) demonstrate how
caching can accelerate data science experiments while maintaining FAIR
principles~\cite{schubotz2022caching}. Their work emphasizes content-addressable
storage for automatic result reuse, similar to the fingerprinting approach
used in this thesis.

However, existing academic work on ML caching does not specifically address
the requirements of cross-validation experiments, where cache validity depends
on explicit encoding of data splits. The challenge of fold-aware cache key
generation remains unaddressed in the literature.

\begin{todobox}
TODO Lennart: Only 2 papers (Krishna 2025, Schubotz 2022) are currently cited
in this section. Find and add 3--5 more ML caching papers. Look for:
\begin{itemize}
\item Pipeline caching in ML workflows (e.g., papers on DVC internals,
      ZenML caching design, or MLflow artifact management)
\item Experiment result reuse and memoization in scientific computing
\item Computational artifact management for iterative ML
\item Content-addressable storage for ML pipelines
\end{itemize}
Search terms: ``ML experiment caching'', ``pipeline artifact reuse'',
``memoization machine learning'', ``computational reproducibility caching''.
\end{todobox}

% ============================================================================
\section{Commercial MLOps Tools}
% ============================================================================
\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Should we incroporate this too here ? Working Draft Note (SotA - The "Gap" Argument):} \\
\textbf{1. The Workflow Orchestration (C3) Gap:} Kreuzberger lists tools like Airflow, Kubeflow, and AWS SageMaker. \textbf{Critique:} These tools are designed for linear pipelines. Argue that they lack "deterministic sub-task invalidation" for nested loops like LOSO (128 subjects $\times$ 9 configs). \\
\textbf{2. Cost Management (P6):} The paper says CT (Continuous Training) frequency should consider costs. \textbf{My Argument:} I am evolving P6 from "periodic retraining" to "state-aware retraining," which is more efficient for research. \\
\textbf{3. Tool Review Weakness:} Reference their "Table I" (Tool Review). Note that most tools focus on the "Production/Serving" end (C8/C9). My framework fills a gap in the "Experimental/Engineering" stage by providing a "pre-computation check."
\end{minipage}
}
\end{center}
Current MLOps platforms provide general experiment tracking and pipeline
caching but are agnostic to the internal structure of cross-validation. While these tools can be configured for LOSO via complex templating, they do not natively encode fold-identity into cache keys, leading to coarse-grained invalidation where changing a single parameter might trigger an entire 128-fold recomputation unnecessarily.

\begin{itemize}
\item \textbf{MLflow:} Provides experiment tracking, model registry, and
      basic result caching~\cite{zaharia2018mlflow}. Pipeline caching requires
      the MLflow Recipes module, which caches individual steps but does not
      natively encode fold assignments in cache keys.

\item \textbf{DVC:} Uses MD5 hashes of pipeline inputs and code to enable
      pipeline stage reuse~\cite{dvc_pipelines}. While DVC can cache
      cross-validation results, it requires manual configuration of the pipeline
      definition to parameterise fold indices---a process not documented in
      standard DVC workflows.

\item \textbf{ZenML:} Implements hash-based step caching with automatic
      invalidation when inputs or code change. However, ZenML's caching
      abstractions do not provide explicit support for encoding held-out
      subject identifiers, requiring users to implement custom materializers.

\item \textbf{Neptune AI and Weights \& Biases:} While these platforms excel at 
      metric visualization and experiment logging~\cite{wandb2024docs}, 
      they are inherently subject to platform stability and vendor lock-in. 
      The discontinuation of Neptune AI services in early 2026~\cite{neptune2026shutdown} 
      vividly underscores the long-term risks of relying on proprietary, cloud-based 
      metadata stores for academic research. Furthermore, these tools primarily 
      function as observation layers; they lack integrated computational caching 
      mechanisms. This necessitates the development of the independent, 
      self-hosted framework for reproducible LOSO-fold management proposed 
      in this thesis.
\end{itemize}

% ============================================================================
\section{Research Gap Analysis}
% ============================================================================
\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Working Draft Note (3.3):}
Hier die Brücke zwischen Prozess-Theorie und technischem Defizit schlagen: Trotz der weiten Verbreitung von CRISP-DM, wie in der Meta-Analyse von \cite{schroer2021crispdmslr} bestätigt, fehlen dem Modell kritische Mechanismen für moderne ML-Workflows. Insbesondere die von \cite{studer2021crispmlq} geforderte Qualitätssicherung (CRISP-ML(Q)) bleibt oft auf einer rein prozessualen Ebene stehen. Hier ansetzen: Mein Framework füllt diese Lücke technologisch, indem es die \textit{Reproduzierbarkeit} nicht nur als Richtlinie sieht, sondern durch striktes \textit{Configuration Fingerprinting} und \textit{Caching} erzwingt, um die in \cite{googlemlops2023} kritisierten manuellen Fehlerquellen zu eliminieren.
\end{minipage}
}
\end{center}

Table~\ref{tab:related_works_comparison} compares caching capabilities of
existing tools relevant to this thesis.

\begin{todobox}
TODO Lennart: Verify table contents against latest tool versions before
final submission. DVC, ZenML, and MLflow update frequently.
\end{todobox}

\begin{table}[htbp]
\centering
\caption{Comparison of caching capabilities in MLOps tools. ``LOSO Support''
indicates whether the tool natively encodes fold assignments in cache keys
or requires manual configuration.}
\label{tab:related_works_comparison}
\begin{tabular}{llll}
\toprule
\textbf{Tool/Framework} & \textbf{Caching Strategy} & \textbf{Fingerprinting} & \textbf{LOSO Support} \\
\midrule
MLflow        & Step-level (optional)          & Parameter hashing          & Manual fold param. \\
DVC           & Pipeline stage                 & MD5 (code + inputs)        & Manual pipeline \\
ZenML         & Step-level                     & SHA-256 (inputs+code+env)  & Custom materializers \\
Neptune / W\&B & None (tracking only)          & N/A                        & N/A \\
\textbf{This Thesis} & \textbf{Hierarchical (feat.+models)} & \textbf{SHA-256 + subject ID} & \textbf{Native fold-aware} \\
\bottomrule
\end{tabular}
\end{table}

The core gap is the lack of validated, documented patterns for
implementing fold-aware caching in cross-validation workflows. While tools
like DVC and ZenML provide general pipeline caching infrastructure, adapting
them for LOSO requires custom logic that:
\begin{enumerate}
\item Encodes the held-out subject identifier in cache keys to prevent
      cross-fold contamination
\item Manages hierarchical dependencies between preprocessing, feature
      extraction, and model training stages
\item Validates cache hits against both parameter configurations and
      fold assignments
\end{enumerate}

These requirements are not addressed in standard documentation, and incorrect
implementation risks subtle bugs where models trained on Subject~A are
erroneously reused for evaluating Subject~B.

This thesis provides a reference implementation demonstrating these patterns
in a real LOSO scenario (128-fold cross-validation on the BOAS dataset), with
explicit measurement of cache performance (hit rates, time savings) and
validation of reproducibility. The contribution is not a new MLOps platform,
but a validated approach for extending existing fingerprinting
techniques to cross-validation contexts.

% ============================================================================
\section{Sleep Stage Classification Methods}
\label{sec:related_sleep}
% ============================================================================

\begin{missingref}
TODO Lennart: Find additional references for this section before final submission:
\begin{itemize}
\item 2--3 feature-based sleep classification papers (2020--2026) beyond
      Satapathy~(2024). Search: ``EEG feature extraction sleep staging'',
      ``handcrafted features polysomnography classification''.
\item 1 LOSO methodology paper for EEG/BCI (e.g., Jayaram et al.\ 2016
      ``Transfer learning in brain-computer interfaces'' or similar).
\item 1 class imbalance handling reference for sleep staging (N1 underrepresentation).
      Search: ``class imbalance sleep staging'', ``macro F1 sleep classification''.
\end{itemize}
\end{missingref}

Sleep stage classification is a well-studied problem in biomedical signal
processing. Methods are broadly grouped into two categories: feature-based
approaches that extract handcrafted signal descriptors, and deep learning
approaches that learn representations directly from raw signals.

\subsection{Feature-Based Approaches}

Feature-based sleep staging extracts domain-informed descriptors from EEG
signals and feeds them to classical machine learning classifiers. Features
typically span time-domain statistics (mean, variance, RMS), frequency-domain
band powers (delta, theta, alpha, sigma, beta), and signal complexity measures
(Hjorth parameters, Hurst exponent, sample entropy).

Satapathy et al.\ (2024) demonstrate that multi-modality feature sets combining
EEG, EOG, and EMG signals with ensemble classifiers achieve competitive
performance on standard sleep staging benchmarks~\cite{satapathy2024multimodal}.
Their work highlights that carefully engineered features remain competitive with
deep learning methods, particularly in cross-subject generalisation settings
where training data is limited per subject.

The approach taken in this thesis follows this feature-based paradigm: 149
handcrafted features are extracted per 30-second epoch across 6 EEG channels,
then classified using XGBoost and Random Forest. Feature-based approaches have
the advantage of interpretability (feature importance scores identify which EEG
descriptors drive classification) and compatibility with tabular ML models that
are well-suited to caching (compact serialised representations).

LOSO cross-validation is the standard evaluation protocol in this domain,
reflecting the fundamental challenge of subject variability: EEG patterns
differ substantially between individuals, and a model trained on one subject
pool may not generalise to unseen subjects. LOSO evaluation provides an
unbiased estimate of cross-subject performance.

\subsection{Deep Learning Approaches}
\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Working Draft Note:} 
wir nutzen kein deep learning wir haben cnn fnn gestrichen sollten wir diese section so lassen? 
\end{minipage}
}
\end{center}
\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Working Draft Note:} 
sollten wir das einbauen """While deep learning architectures such as CNN-LSTMs can achieve high accuracy, recent comparative analyses by Siddiqui et al. (2025) confirm that these models require "substantially more computational resources" compared to traditional ensembles~\cite{siddiqui2025}. Their findings show that Random Forest classifiers remain highly competitive, offering a superior balance between interpretability and computational efficiency. This trade-off is critical for the framework proposed here; while a deep model like U-Sleep~\cite{perslev2021usleep} is difficult to cache due to its massive parameter size and non-deterministic training times, the feature-based XGBoost and Random Forest models used in this work are ideal candidates for granular, fold-level caching. By prioritizing these "caching-friendly" models, this work enables a rapid hyperparameter search across the entire LOSO spectrum that would be computationally prohibitive with deep learning alternatives."""

\end{minipage}
}
\end{center}
Deep learning methods for sleep staging learn feature representations directly
from raw EEG signals, eliminating the need for manual feature engineering.

Perslev et al.\ (2021) present U-Sleep, a fully convolutional network trained
on diverse polysomnographic recordings that achieves robust high-frequency sleep
staging across different acquisition systems~\cite{perslev2021usleep}. U-Sleep
demonstrates that deep models can generalise across hardware configurations, a
significant practical advantage.


\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Working Draft Note:} 
i want to remove jirakittayakorn2024zleepanlyst feels irrelevant also i am not using cnn rnn or dl so this entire subsesction feels a bit too long)
\end{minipage}
}
\end{center}
Jirakittayakorn et al.\ (2024) propose ZleepAnlystNet, a deep learning model
for automatic sleep staging from single-channel raw EEG using separating
training strategies~\cite{jirakittayakorn2024zleepanlyst}. Single-channel
approaches are particularly relevant for wearable sleep monitoring applications
where full PSG electrode montages are impractical.

Zhang et al.\ (2025) address domain generalisation in sleep staging through an
adversarial domain-generalised residual attention network, improving robustness
to dataset shift across recording sites and equipment~\cite{zhang2025adg}.

Li et al.\ (2025) explore few-shot sleep classification using spatial-temporal
hypergraph meta-learning, targeting scenarios with limited labelled
data~\cite{li2025metasth}.

A common limitation of deep learning approaches in the context of this thesis
is their computational and caching profile: deep models (CNNs, RNNs,
Transformers) typically serialise to hundreds of MB or GB of parameters, and
their training time is dominated by GPU availability rather than CPU caching
bottlenecks. None of the cited deep learning papers address caching strategies
for iterative experimentation or LOSO efficiency. This gap motivates the
caching-oriented framework developed here.

\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Working Draft Note:} 
sollten wir "\paragraph{Computational and Caching Profile.} 
A common limitation of these deep learning approaches in the context of this thesis is their specific resource profile. While models like \textit{ADG-RANet} \cite{zhang2025adg} and \textit{MetaSTH-Sleep} \cite{li2025metasth} push the boundaries of domain generalization and few-shot adaptation, they do so at a significant computational cost. These architectures typically serialize to hundreds of megabytes, and their training is dominated by heavy GPU utilization. 

As noted by \cite{siddiqui2025}, traditional ensemble methods like Random Forest remain highly competitive, offering a superior balance between interpretability and resource efficiency. This trade-off is central to the framework proposed here: while the deep learning alternatives cited above do not address caching strategies for iterative LOSO experimentation, the feature-based models used in this work are ideal candidates for granular, fingerprint-based caching. By prioritizing these "caching-friendly" models, this work enables a rapid hyperparameter search that would be computationally prohibitive with deep learning alternatives." adden 

\end{minipage}
}
\end{center}

\subsection{LOSO Evaluation in EEG Research}

Leave-One-Subject-Out cross-validation is preferred over standard k-fold
cross-validation in EEG and sleep research for a fundamental reason: EEG
signals contain strong subject-specific signatures. If training and test data
from the same subject appear in the same fold (as in random k-fold splits),
the classifier learns subject-specific patterns rather than generalisable
sleep stage indicators. LOSO ensures the model is evaluated on data from
subjects it has never seen, providing a conservative and clinically meaningful
estimate of generalisation.

The computational cost of LOSO scales directly with subject pool size. For
128 subjects, LOSO requires 128 separate model training runs per configuration.
With 18 configurations, this yields 2{,}304 training operations per complete
experiment---making caching not merely convenient but essential for practical
parameter search.

% ============================================================================
\section{Industrial Research and Applications}
\label{sec:related_industrial}
% ============================================================================

\begin{todobox}
TODO Lennart: The professor explicitly requires an industrial research section.
Find 2--3 references on industrial ML deployment or commercial sleep technology:
\begin{itemize}
\item Industrial MLOps / production ML systems (e.g., Google's ML production
      paper, Facebook/Meta's production ML infrastructure)
\item Commercial sleep monitoring devices (Dreem headband, Fitbit sleep staging,
      Apple Watch sleep tracking accuracy papers)
\item Industry-academic gap in sleep staging deployment
\end{itemize}
Search: ``production machine learning systems'', ``commercial sleep staging device'',
``wearable EEG sleep monitoring accuracy''.
\end{todobox}

The gap between academic sleep staging research and industrial deployment
reflects broader challenges in production machine learning. Academic studies
typically report results on controlled datasets under ideal acquisition
conditions, while commercial sleep monitoring must cope with consumer-grade
hardware, motion artefacts, and users who do not follow electrode placement
protocols.

\textbf{Commercial Sleep Monitoring:} Consumer wearables (smartwatches,
fitness bands) incorporate simplified sleep staging using accelerometers and
optical heart rate sensors rather than full EEG. While convenient, these
devices sacrifice staging accuracy, particularly for distinguishing N1 from
N2 and accurately identifying REM. Medical-grade consumer EEG headbands
(such as the Dreem headband) have demonstrated PSG-comparable performance
in controlled studies, suggesting that the accuracy gap is closeable with
sufficient EEG channel coverage.

\textbf{Production ML Challenges:} Deploying ML models in clinical and
consumer contexts introduces requirements not typically addressed in academic
research: model versioning, inference latency constraints, regulatory approval,
and the need for reproducible audit trails. Caching and fingerprinting
frameworks like the one developed in this thesis address the reproducibility
and computational efficiency aspects of this deployment challenge, particularly
for iterative model development and validation cycles.

\textbf{Research-to-Production Gap:} The transition from research prototype
to production deployment requires systematic experiment management, including
the ability to reproduce any prior result on demand and to resume interrupted
experiments without restarting from scratch. These requirements directly
motivate the fingerprint-based caching app
\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Working Draft Note (3.5):}
Diskussion der Diskrepanz zwischen akademischem Anspruch und industrieller Realität. Während \cite{studer2021crispmlq} mit CRISP-ML(Q) ein theoretisches Prozessmodell für Qualitätssicherung vorgibt, definieren \cite{kreuzberger2023mlops} und \cite{googlemlops2023} die notwendige Architektur. Mein Framework schließt die Lücke in der \textit{Monitoring & Maintenance}-Phase von Studer, indem es die technische Reproduzierbarkeit auf Hardware-Ebene (trotz GPU-Varianz) durch striktes Metadaten-Tracking sicherstellt.
\end{minipage}
}
\end{center}
