\chapter{Summary and Future Work}
\label{chap:summary}


% ============================================================================
\section{Summary}
% ============================================================================

This thesis developed a fingerprint-based caching framework for
Leave-One-Subject-Out (LOSO) cross-validation experiments. The core idea is
simple: each experiment configuration is mapped to a deterministic SHA-256
hash computed from the canonical JSON representation of the tracked experiment
state---random seed, code version, model type and hyperparameters, feature
configuration, and held-out subject identifier. The resulting fingerprint
serves simultaneously as cache key, reproducibility identifier, and
experiment label.

The inclusion of the held-out subject ID in the fingerprint is the critical
design decision that distinguishes this framework from generic pipeline
caching. By encoding the LOSO fold identity in the cache key, the system
prevents cross-fold contamination: a model trained with Subject~$S_i$ held
out cannot be retrieved when Subject~$S_j$ is the test subject, because the
fingerprints differ. This guarantee is enforced structurally, not by
convention, eliminating a class of subtle caching bugs that could
compromise scientific validity.

The framework was validated through a sleep stage classification case study
using the BOAS dataset (128 subjects, approximately 120{,}000 epochs of
30-second EEG recordings across 6 channels). The experimental evaluation
comprised 18 configurations (2 models $\times$ 3 correlation thresholds
$\times$ 3 feature counts) with 128-fold LOSO cross-validation, totalling
2{,}304 model training operations.

\textbf{Key findings:}

\begin{enumerate}
\item \textbf{Substantial speedups for cache-viable models.}
      XGBoost achieved 37.5--202$\times$ speedup (median 54$\times$) with
      compact cache sizes of approximately 185~MB. The best configuration
      (all 149 features, no correlation filtering) reduced a 3{,}816-second
      cold run to 18.9~seconds. Random Forest achieved 11--23.5$\times$
      speedup but required approximately 18.5~GB of cache storage, making
      it borderline viable.

\item \textbf{Best classification accuracy of 85.5\%.}
      XGBoost with all 149 features and no correlation filtering achieved
      the highest LOSO accuracy (85.5\%, Cohen's $\kappa = 0.737$,
      macro F1 = 0.636), demonstrating that the caching framework does not
      compromise classification performance.

\item \textbf{Clear model viability spectrum.}
      A 15-model viability study identified 11 cache-viable models
      (including all boosting methods and SVMs) and 4 non-viable model
      types (Random Forest, Extra Trees, KNN $k=5$, KNN $k=10$) whose
      large serialised representations negate caching benefits. Viability
      verdicts (VIABLE/NOT\_VIABLE) were consistent across two hardware
      configurations; absolute speedup ratios were not, since faster CPUs
      compress cold training time without proportionally reducing I/O-bound
      warm load time.

\item \textbf{Cache benefit scales with dataset size.}
      SVM-RBF speedup increased from 9$\times$ at 10 subjects to
      201$\times$ at 128 subjects. Training cost grows quadratically with
      dataset size while cache loading cost remains constant, meaning
      caching becomes more valuable as datasets grow.

\item \textbf{100\% reproducibility.}
      All cached models produced bitwise-identical predictions to cold
      retraining across all 18 configurations and 128 LOSO folds (verified
      via \texttt{numpy.array\_equal()}).
\end{enumerate}

\subsection{Research Questions Revisited}

The four research questions posed in Chapter~1 are answered below. Where the
answer is affirmative, the nuance matters as much as the confirmation.

\begin{itemize}
\item \textbf{RQ1 (Efficiency):} Confirmed, but efficiency is model-dependent.
      The framework delivers order-of-magnitude speedups for compact models
      (XGBoost median 54$\times$, peak 202$\times$) and the persistent
      Layer~1 feature store eliminates redundant extraction entirely. However,
      the speedup ratio is not a fixed property of the framework --- it
      reflects the ratio of training time to serialization size, and depends
      on hardware: ranges from $\sim$9.5$\times$ (RF on a fast desktop) to
      over 18{,}000$\times$ for the most compact boosting models at scale.
      Efficiency is the framework's primary claim; it is real, but the
      magnitude depends on model choice and hardware speed.

\item \textbf{RQ2 (Reproducibility):} Confirmed unconditionally for
      deterministic estimators. Identical configurations produce identical
      SHA-256 fingerprints, and cached models produce bitwise-identical
      predictions across all 18 configurations and 128 LOSO folds. The
      key design insight --- encoding the held-out subject ID in the cache
      key --- enforces fold isolation structurally rather than by convention,
      eliminating a class of subtle scientific validity bugs without any
      performance cost.

\item \textbf{RQ3 (Scalability):} Confirmed directionally. Cache benefit
      increases with dataset size for models whose training complexity scales
      super-linearly (SVM-RBF: 9$\times$ at 10 subjects to 201$\times$ at 128
      subjects). For linearly-scaling models the speedup is approximately
      stable but already high (SVM-Linear: $\sim$2{,}000$\times$ across all
      scales). The implication is that caching becomes \textit{more} valuable
      as studies grow, not less. A second counterintuitive result from
      cross-machine comparison: caching also becomes more valuable on
      \textit{slower} hardware. Cache load time is I/O-bound and approximately
      constant across machines, while cold training time is CPU-bound and
      scales with hardware speed. A researcher on a standard laptop who trains
      $3\times$ slower than a high-end workstation observes speedup ratios
      $\sim$$3\times$ larger --- the framework delivers its greatest gains
      precisely where compute resources are most constrained.

\item \textbf{RQ4 (Trade-offs):} A clear model viability spectrum exists, but
      the VIABLE/NOT\_VIABLE boundary is not universal --- it depends on the
      available hardware and how many re-runs are planned. 11 of 15 evaluated
      model types are unambiguously viable regardless of context. The 4
      not-viable types (Random Forest, Extra Trees, KNN) fail because their
      serialized representations (130--1{,}544~MB per fold) make loading
      approach retraining time. Random Forest occupies a borderline zone: at
      11.6$\times$ median speedup and 18.5~GB per run, one re-run saves up to
      $\sim$2~hours per configuration at substantial storage cost --- worthwhile
      for some workflows, not others.
\end{itemize}

The contribution of this thesis is not a new MLOps platform, but a validated
approach for extending fingerprinting techniques to cross-validation contexts.
All findings reported here are conditional: speedup ratios depend on hardware,
viability verdicts depend on the threshold a researcher sets, and accuracy
results are specific to the BOAS EEG dataset and the 149-feature pipeline.
What the framework provides is not a universal answer but a \textit{methodology
for arriving at the answer} for a given experimental context --- the fingerprint
mechanism, the viability measurement approach, and the two-layer cache design
are transferable; the specific numbers are not.

\subsection{Practical Takeaways}

Three concrete guidelines emerge from this work for researchers considering
fingerprint-based LOSO caching.

\textbf{First, check the cache efficiency metric before committing to a full run.}
The $\eta$ metric (MB stored per second saved, Section~\ref{sec:evaluation_plan})
can be estimated from a short pilot (e.g.\ 5 folds) without running the full
128-fold experiment. A pilot that yields $\eta < 0.5$ reliably predicts a
full-scale speedup above 10$\times$; a pilot above 2.0 signals that warm
runs will not recover meaningful time regardless of dataset size.

\textbf{Second, the held-out subject ID must be part of the cache key.}
Omitting it is the single most dangerous simplification a developer might make
when adapting this framework to a new context. The structural enforcement
provided by the fingerprint prevents a subtle correctness bug where models
trained on overlapping data are silently returned as fold-specific results.
No test catches this failure without inspecting individual fold compositions.

\textbf{Third, budget for the cold run on slower hardware.}
The full 18-configuration cold run required approximately 78 hours on a
standard desktop (PC1) versus 22 hours on a high-end workstation (RTX 5090).
On resource-constrained hardware, a staged approach---running configurations
in batches across sessions---is practical because the framework guarantees
that completed folds are never recomputed between sessions.

\subsection{Limitations and Scope}

The results reported in this thesis are bounded by three scope decisions.
First, both model families evaluated at full scale (XGBoost, Random Forest)
are ensemble tree methods; neural network models were excluded due to
environment constraints. Second, the classification target (five-class AASM
sleep staging) is moderately imbalanced and domain-specific; results may differ
on tasks with more extreme imbalance or very high-dimensional inputs.
Third, the global ANOVA feature selection introduces $\approx 0.8\%$ label
leakage per fold (Section~\ref{sec:feature_selection_strategy}); the
empirically measured accuracy impact was $\leq 1.1\%$, which is within
inter-run variability for this dataset but represents a methodological bound
on the accuracy estimates reported.

These limitations do not undermine the primary thesis claim --- that
fingerprint-based LOSO caching is a viable, correct, and reproducible approach
--- but they define the conditions under which the specific numerical results
hold.


% ============================================================================
\section{Future Work}
% ============================================================================

Several directions could extend the framework developed in this thesis.

\subsection{Lazy Per-Fold Feature Selection Cache}
\label{sec:future_work}

The global ANOVA feature selection used in this thesis introduces minor
information leakage: each LOSO fold's held-out subject contributes
$1/128 \approx 0.8\%$ of the data used to fit the global feature ranking.
A scientifically correct implementation would fit ANOVA independently within
each fold on the 127 training subjects only.
Na\"ively, this requires re-running ANOVA on every warm run before the cache
key can even be computed --- because the current fingerprint includes the
selected feature names, which are only known after ANOVA has run.
Our benchmark (Figure~\ref{fig:global_vs_perfold}) quantifies the consequence:
without a fix, the XGBoost warm-run speedup degrades from 22.9$\times$
(global) to 4.4$\times$ (na\"ive per-fold) at $N=20$ subjects, because ANOVA
re-runs 20 times even on a full cache hit.

A \textit{lazy selection cache} resolves this without any accuracy cost.
ANOVA (\texttt{f\_classif}) is a fully deterministic computation: given the
same training subjects and selection parameters, it always produces the same
ranked feature list. Two implementation variants are conceivable.

\textbf{Version A --- Determinism-based key (recommended for static datasets).}
Remove the selected feature names from the model-cache fingerprint and
introduce a lightweight \emph{selection registry} keyed by experiment
configuration and fold identity:
%
\begin{equation*}
  k_{\text{sel}} = \text{SHA-256}\!\left(
    \textit{dataset\_version} \;\Vert\; \textit{code\_version} \;\Vert\;
    \textit{corr} \;\Vert\; \textit{top\_k} \;\Vert\; \textit{method}
    \;\Vert\; \textit{held\_out\_subject}
  \right)
\end{equation*}
%
The key must include \texttt{dataset\_version} and \texttt{code\_version} so
that changes to the dataset (e.g.\ a subject removed or corrected) or the
preprocessing code correctly invalidate the registry. Without these fields,
the same key could return a stale feature list after the underlying data
changes --- a silent correctness bug. On a cold run, ANOVA is fitted and the
sorted selected feature names are stored under $k_{\text{sel}}$.
On a warm run, the registry is queried first (${\sim}1$~ms); the feature list
is retrieved, the full model fingerprint is assembled, and the cached model is
loaded without running ANOVA at all.
This restores warm-run time to ${\approx}7$~s (registry lookup + model load).
The cold-run overhead is $128 \times 22~\text{ms} \approx 2.8~\text{s}$.
Version A is appropriate for a frozen dataset such as BOAS where the
held-out subject uniquely identifies the training fold.

\textbf{Version B --- Data-aware key (recommended for production).}
Add an explicit training-set fingerprint to $k_{\text{sel}}$, for example a
hash of the sorted training subject IDs or a checksum of the feature matrix.
This binds the registry entry unambiguously to a specific data snapshot and
protects against stale cache hits if subjects are later removed, preprocessing
changes, or fold construction is modified. Version B is the production-grade
design; Version A is a valid shortcut for immutable datasets only.
Neither variant was implemented in this thesis --- both are designs for
future work.

In summary, a global-scope fast-screening run (current design) and a per-fold
final-reporting run (Version A or B registry) can coexist in the same framework.
The global run uses a single fingerprint shared by all folds; the per-fold run
adds one registry lookup per fold and eliminates the residual $0.8\%$ leakage
by construction.

\subsection{Scaling to Real-World Use}

Two directions would move the framework from research prototype to deployable
tool.

\textbf{Cross-domain validation.}
The framework was validated on EEG sleep classification. The underlying
fingerprinting approach is domain-agnostic, and applying it to other domains
with subject-level cross-validation --- brain-computer interfaces,
speaker-independent speech recognition, patient-level medical imaging splits
--- would test generalisability and surface domain-specific requirements not
covered by the current implementation.

\textbf{Integration with MLOps platforms.}
The framework could be packaged as plugins for DVC, ZenML, or MLflow. A DVC
plugin could extend the pipeline definition language to natively parameterise
LOSO fold identifiers; a ZenML custom materialiser could encode held-out
subject IDs in step cache keys. Such integrations would make fold-aware
caching accessible to researchers who already use these platforms without
requiring a custom framework. The viability verdicts established in this thesis
(which model families are and are not worth caching, at what storage cost)
would directly inform the default caching policy of such plugins.

\subsection{Production Deployment Considerations}

Extending to production environments would require selective caching policies
that prioritise high-reuse artefacts, automatic eviction based on access
frequency and storage budgets, and hierarchical invalidation where upstream
changes (e.g.\ preprocessing updates) propagate to all dependent caches.
A shared cache repository would also allow Layer~1 feature stores to be
reused across experiments that share the same preprocessing pipeline but
differ in model hyperparameters.

A practical benefit not captured by the speedup metric is thermal load
reduction on constrained hardware. Cold runs can be interleaved across
sessions --- running a batch of configurations, allowing the machine to cool,
resuming later --- with the guarantee that completed folds are never
recomputed. The 9-hour RF cold run is the clearest illustration: splitting
it across three sessions is trivially safe under the fingerprint guarantee.

\subsection*{A Research Trajectory}

This thesis is one step in a longer research arc.
At the bachelor level, the contribution is an empirical viability map:
which model families are worth caching under LOSO, at what storage cost,
and under what conditions the benefit scales. The fingerprinting mechanism
and two-layer cache design are validated but not automated.

A natural master-level extension would exhaustively characterise viability
across domains, model families, and dataset scales --- building a
\textit{viability knowledge base} that captures not just which models cache
well on BOAS, but the generalised conditions under which any model becomes
cache-viable or not.

At the PhD or applied research level, that knowledge base could power an
\textit{intelligent caching orchestrator}: a system that accepts a researcher's
experimental plan (models, hyperparameter grid, dataset, hardware), queries the
viability knowledge base, and automatically configures a caching strategy ---
which layers to cache, what storage budget to allocate, when to evict, and
how to schedule cold runs to avoid thermal constraints. Paired with an
interactive planning interface (the researcher declares their workflow goals;
the system recommends a caching configuration before the first run begins),
this would make fold-aware caching a zero-configuration tool rather than a
framework a researcher must design themselves. The building blocks are already
available --- this thesis provides the empirical foundation; the referenced
MLOps platforms provide the deployment infrastructure. What remains is the
intelligence that connects them.
