\documentclass{imc-inf} % Use the actual IMC template class

% --- IMC Template Specific Commands ---
% These commands are defined by the imc-inf.cls file
\title{Optimizing ML Experiments: Intelligent Frameworks for Efficiency and Reproducibility}
\subtitle{A Fingerprint-Based LOSO Caching Approach Validated on EEG Sleep Classification}
\thesistype{Bachelor Thesis} % FIXED: Changed from Bachelor Exposé
\author{Lennart Gorzel}
\supervisor{Prof. Himanshu Buckchash} % FIXED: Correct supervisor
\copyrightyear{2026} % FIXED: Updated year
\submissiondate{05.05.2026} % FIXED: Proposal deadline
\keywords{Machine Learning, Intelligent Caching, Reproducibility, Sleep Stage Classification, EEG, Feature Engineering}

% --- Standard Packages (keep these) ---
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{svg}
\usepackage{rotating}
\usepackage{afterpage}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{booktabs} % For professional tables
\usepackage{multirow} % For merged cells in tables
\usepackage{xcolor}

% Define the missing plotsuggestion environment
\newenvironment{plotsuggestion}{%
    \begin{center}
    \begin{minipage}{0.9\textwidth}
    \begin{itshape}
    \textbf{Plot Suggestion:} 
}{%
    \end{itshape}
    \end{minipage}
    \end{center}
}

\begin{document}

% --- Front Matter (as per IMC template) ---
\frontmatter
\maketitle

\begin{declarations}
\end{declarations}



% ============================================================================
% ABSTRACT
% ============================================================================

\begin{abstract}
This thesis addresses the computational costs and reproducibility challenges inherent in modern Machine Learning (ML) experimentation. As ML systems increase in complexity, researchers face significant overhead from redundant computations, inconsistent results across runs, and difficulties resuming interrupted experiments. This work proposes a fingerprint-based caching framework utilizing SHA-256 configuration hashing to systematically mitigate these challenges.

The research develops a deterministic fingerprinting system where a cryptographic hash uniquely identifies each experiment configuration. The fingerprint is derived from a canonical JSON representation of the tracked experiment state, including the random seed, code version, model hyperparameters, feature configuration, and held-out subject. This ensures that any modification to parameters that may affect the output results in a unique fingerprint, enabling automatic cache invalidation. Consequently, identical fingerprints indicate an identical tracked configuration, allowing for valid, reproducible cache retrieval.

The framework specifically extends this mechanism to Leave-One-Subject-Out (LOSO) cross-validation. By incorporating the held-out subject into the cache key, the system ensures cache integrity and prevents the improper reuse of models across different folds, thereby maintaining strict isolation between training and testing sets under a fixed dataset partition.

Validation is conducted via a sleep stage classification case study using the Bitbrain Open Access Sleep (BOAS) dataset (128 subjects, approximately 120,000 epochs). The feature extraction module computes 149 features per epoch across six EEG channels, which are cached persistently to isolate the performance evaluation to the computationally intensive model training phase. 

The contributions of this thesis are: (1) a fingerprint-based LOSO model caching framework that ensures cache correctness and prevents cross-fold contamination; (2) quantified efficiency gains, demonstrating a median speedup of 178x for XGBoost and 17.7x for Random Forest across 128 LOSO folds; and (3) validated reproducibility, where all cached models yielded identical classification results to those obtained through cold retraining.
\end{abstract}

% 3. THE AI TABLE & THANKS to HImanshu (This is where the Acknowledgements go!)
\begin{acknowledgements}
    I would like to express my sincere gratitude to my supervisor, Prof.\ Himanshu Buckchash, for his continuous guidance, constructive feedback, and patience throughout the development of this thesis. His expertise in machine learning and his emphasis on methodological rigour shaped both the research direction and my understanding of scientific work.

    I also wish to thank the IMC University of Applied Sciences Krems for providing the academic environment and resources that made this research possible. The structured curriculum in the Computer Science programme laid the foundation for the interdisciplinary skills required in this project.

    Finally, I would like to thank my family and friends for their support and encouragement during the writing of this thesis.

    \vspace{1cm}
    \section*{AI Usage Disclosure (per FHR-5-0044)}
    \begin{center}
    \colorbox{yellow!20}{%
    \begin{minipage}{0.95\textwidth}
    \small
    In accordance with the IMC examination regulations, the following AI-assisted tools were used during the preparation of this thesis. In all cases, the underlying research, analysis, and scientific argumentation were performed independently by the author. AI tools were used exclusively in a supporting capacity as permitted under FHR-5-0044. Original texts and prompts are available upon request.

    \vspace{0.3cm}
    \begin{tabular}{|p{0.18\textwidth}|p{0.25\textwidth}|p{0.25\textwidth}|p{0.22\textwidth}|}
        \hline
        \textbf{KI-Tool} & \textbf{Wie?} & \textbf{Warum?} & \textbf{An welcher Stelle?} \\ \hline
        DeepL & Translation and linguistic polishing of self-written German drafts into English & Grammar and fluency improvement of author-written text & All chapters \\ \hline
        GitHub Copilot & Code completion suggestions during implementation & Accelerating routine coding tasks (boilerplate, syntax) & Python caching framework (Ch.~4) \\ \hline
        Claude (Anthropic) & Code refactoring assistance and debugging support & Identifying bugs in caching logic; restructuring existing code & Python pipeline code (Ch.~4) \\ \hline
        Gemini (Google) & \LaTeX{} template debugging and formatting assistance & Resolving compilation errors in document formatting & Document template \\ \hline
    \end{tabular}
    \end{minipage}}
    \end{center}
\end{acknowledgements}

\addtoToC{Table of Contents}%
\tableofcontents%
\clearpage

\addtoToC{List of Tables}%
\listoftables
\clearpage

\addtoToC{List of Figures}%
\listoffigures
\clearpage

% ============================================================================
% MAIN MATTER
% ============================================================================
\mainmatter

% Chapters
\include{chapters/01_introduction}
\include{chapters/02_background}
\include{chapters/03_related_work}
\include{chapters/04_methodology}
\include{chapters/05_summary}

% ============================================================================
% BACK MATTER
% ============================================================================
\backmatter%

\begin{center}
\colorbox{yellow!20}{%
\begin{minipage}{0.9\textwidth}
\textbf{Working Draft Note:} 
TODO Lennart: Currently $\sim$18 references are cited (30 in bib file, but
12 are uncited). Professor expects a minimum of 25 cited references. Missing
categories:

\begin{itemize}
\item BOAS dataset paper (replace \texttt{kemp2000sleepEDF} / \texttt{goldberger2000physionet})
\item 2--3 feature-based sleep classification papers (see Ch3 \texttt{missingref})
\item 1 LOSO methodology paper for EEG/BCI
\item 1 ANOVA / feature selection reference
\item 1 EEG signal processing textbook (see Ch2 \texttt{missingref})
\item 1 ML textbook: Hastie ESL or Breiman 2001 (see Ch2 \texttt{missingref})
\item 1 class imbalance / macro F1 reference
\item 2--3 industrial / production ML references (see Ch3 \texttt{todobox})
\end{itemize}
See colored boxes throughout chapters for specific locations where citations
are needed.
\end{minipage}}
\end{center}

\addtoToC{Bibliography}
\bibliographystyle{IEEEtranS}
\bibliography{references}

\end{document}
